Designing and Modeling a Browser-Based Distributed Evolutionary Computation System

### 2015-05-09 GECCO/wk0705.tex
Changing name again

\begin{abstract}
Web browsers have scaled from simple page-rendering engines to operating
systems that include most services the lower OS layer has, with the
added facility that applications can be run by just visiting a web
page. In this paper we will describe the front and back end of a
distributed evolutionary computation system that uses the browser's
capabilities of running programs written in JavaScript. We will focus on two
different aspects of volunteer computing: first, the
pragmatic: where to find those resources, which ones can be used, what
kind of support you have to give them; and then, the theoretical: how
evolutionary algorithms can be adapted to an environment in which nodes
come and go, have different computing capabilities and operate in
complete asynchrony of each other. We will examine the setup needed to
create a  simple distributed evolutionary algorithm using
JavaScript and then find a model of how users react to it by
collecting data from several experiments featuring different classical
benchmark functions. 
\end{abstract}

@@

Most theory and experimentation in parallel computing has been devoted to predict and
optimize the performance in systems where the number of nodes, their
connections, and the time every one is dedicating to the computation is
known in advance as it often did in the Big Science era, where labs
could easily buy 16-node clusters or subscribe to the resources of a
big grid. Even today, you can rent instances of virtual machines in
the cloud relatively cheaply and use them for evolutionary computation
experiments \cite{sherry2012flex}. Nothing beats free, which is one of
the reasons why the era of
Citizen Science already started a few years ago (with SETI@home \cite{david-seti:home} and then BOINC \cite{boinc_grid04}) and it
offers a vast amount of computational resources to be used, if only
you know how to attract them. But there is a challenge: knowing, or at least having a
ballpark, of how your algorithm is going to perform in this uncertain
environment, where none of the factors is known: neither the number of
nodes, through how they are connected, to how long are they going to
be focused on doing what you want them to. That is why some effort is
needed to first understand the dynamics of the decision to participate
in an experiment that requires you to click on a link and then stay
for a while looking at the screen (or just leave it there running).

Besides, since Amazon started selling EC2 several years ago, reliable
and scalable computing resources are also available for a low price
and on demand. Recently, Google has also refurbished its offering
lowering their prices. This means that the conjunction of free or
low-cost cloud computing engines, volunteer computing systems and
untapped capability of desktop systems can be used for creating
massive, or at least potentially massive, distributed computing
experiments. These experiments can be easily created using open-source
repository sites like GitHub\footnote{\url{http://www.github.com}} and deployed automatically to Platform as
a Service (PaaS) sites such as
Heroku\footnote{\url{https://www.heroku.com/}} or
OpenShift\footnote{\url{https://www.openshift.com/}} or even
Infrastructure as a Service (IaaS) sites such as Microsoft's Azure.

@@

Using the web as a resource for distributed, or plainly user based,
evolutionary computation has a long history since JavaScript was
introduced as a browser-based language in 1997 and even before, when
other procedures, including Flash animations, VBScript, ActiveX or Java applets were
used. Java was pointed out in \cite{soares1998get} as a ``language for
internet'', providing some advantages such as multi-architecture compatibility or 
security mechanisms.
%\begin{quote}
%Java can bring some important advantages... solves architectural
%heterogeneity [...] easy to install and [...] built-in security
%mechanisms
%\end{quote} FERGU: quito esta cita y lo pongo sin el quote
% OK, pero ¿por? - JJ
In that work, Soares et al. describe JET, a system that supports
the execution of parallel applications over the Internet. This system has 
a comprehensive statistics support, allowing its use for science, 
unlike the other systems compared in that paper.

@@

The server program includes also additional logic for logging to
different files or services, routes to obtain information from the
server, such as the anonymized IPs that have participated in the
experiment, the ordinal number of the experiment that is being performed or the content of the chromosome pool, and code to serve the static pages
and the JavaScript/CSS
code also. The main changes to the code from the  the previous paper
\cite{arxiv:modeling} is, besides the inclusion of logging for a
better experiment analysis, the function that checks for the algorithm
termination condition and restarts the chromosome pool when
it does; it includes a configurable termination
condition that checks for the solution and empties the chromosome pool
and logs the end of the experiment. This allows us to run continuously
experiments without restarting the server by hand as we did in the
previous version. %FERGU: es esta la única innovación al paper de
                  %Arxiv? A lo mejor no deberíamos mencionarlo ni
                  %usarlo como base
% La principal innovación es que es más fácil recoger datos del
% servidor y que el despliegue es automático - JJ

In this experiment we have used OpenShift (\url{http://openshift.com})
and Heroku (\url{http://heroku.com}) which are Platform as a Service products with a
free tier. It could be also be deployed, with small
or no modification, to other similar PaaS, maybe with a bit of adaptation. Any of them are
free up to a certain level of use and thus fulfill the requirement of
using free resources as a sustainable way of performing distributed
evolutionary computation experiments \cite{Merelo2014,jj:idc:lowcost}.
%
\begin{figure*}[h!tb]
        \centering
        \includegraphics[width=0.9\linewidth]{img/loggly.png}
        \caption{Loggly dashboard showing the evolution of fitness for
        the last 10 minutes of an experiment.}\label{fig:loggly}
\end{figure*}

Programming the server so that it can be deployed, without
modification, to any PaaS is not difficult, but you have to consider
the different capabilities of each one.\begin{itemize}
\item Heroku does not have a permanent filesystem, so if you want to
  create any directory it has to be done when the server is
  started. It is only guaranteed to last while the application is
  running. 
\item In order to scale properly, all requests in Openshift are routed
  through a proxy. This means that the requests all have, as remote
  address, this IP. Other ways of obtaining the requests must be found
  out. On the other hand, it does have a permanent directory for data.
\end{itemize}

Every PaaS has its own way of defining environment variables, which is the way
used for passing private information such as passwords to the
server. This can be done usually through a command-line client ({\tt
  heroku} or {\tt rhc}), which is also used to check the logs or
restart the client. Eventually, we decided to leave the two servers
running.

@@

Some interesting conclusions can be drawn from this data. The first is
not new: we need frameworks able to work with asynchronous algorithms
to take advantage of all the processing power that is available in an
ad-hoc, ephemeral, setup such as the one available in evolutionary
algorithms. The second is that there is going to be CPU available from
volunteers even when no active announcement is made, just from people
that come from a link in a social media page or the Twitter
timeline. However, an announcement brings immediate attention to the
page which will, obviously, depend on the popularity of the social
media and the person announcing it. How this dependency behaves is
still an open issue. Finally it is also an interesting conclusion
that, given enough time and free computing power, we can solve a
difficult optimization problem. Eventually.

\section{Conclusion}
\label{sec:conc}

@@

It also remains to be done a comparison of single-desktop computer
experiment compared with this parallel setup. It is quite clear that,
due to overhead, several volunteers will be needed to match the speed
of a single several-core desktop computer. This will be explored in
future work.

>>>>

\begin{abstract}
Web browsers have scaled from simple page-rendering engines to operating
systems that include most services the lower OS layer has, with the
added facility that applications can be run by just visiting a web
page. In this paper we will describe the front and back end of a
distributed evolutionary computation system that uses the browser's
capabilities of running programs written in JavaScript. We will focus on two
different aspects of volunteer computing: first, the
pragmatic: where to find those resources, which ones can be used, what
kind of support you have to give them; and then, the theoretical: how
evolutionary algorithms can be adapted to an environment in which nodes
come and go, have different computing capabilities and operate in
complete asynchrony of each other. We will examine the setup needed to
create a  simple distributed evolutionary algorithm using
JavaScript, with the intention of eventually  finding a model of how users react to it by
collecting data from several experiments featuring a classical
benchmark function. 
\end{abstract}

@@

Most theory and experimentation in parallel computing has been devoted to predict and
optimize the performance in systems where the number of nodes, their
connections, and the time every one is dedicating to the computation is
known in advance as it often did in the Big Science era, where labs
could easily buy 16-node clusters or subscribe to the resources of a
big grid. Even today, you can rent instances of virtual machines in
the cloud relatively cheaply and use them for evolutionary computation
experiments \cite{sherry2012flex}. Nothing beats free, which is one of
the reasons why the era of
Citizen Science already started a few years ago (with SETI@home \cite{david-seti:home} and then BOINC \cite{boinc_grid04}) and it
offers a vast amount of computational resources to be used, if only
you know how to attract them. But there is a challenge: knowing, or at least having a
ballpark, of how your algorithm is going to perform in this uncertain
environment, where none of the factors is known: neither the number of
nodes, through how they are connected, to how long are they going to
be focused on doing what you want them to. That is why some effort is
needed to first understand the dynamics of the decision to participate
in an experiment that requires you to click on a link and then stay
for a while looking at the screen (or just leave it there
running). Volunteer computing experiments were set up for the first
time in the 90s, in the shape of the SETI@home screensaver
\cite{david-seti:home} that
analyzed signals looking for patterns indicating the existence of
extraterrestrial intelligence due to budget cuts by the US
congress. They proved \cite{anderson2006computational}
that 
\begin{quote}
... [V]olunteer computing can support
applications that are significantly more data-intensive,
or have high memory and storage requirements, than
those in current projects.
\end{quote}
Evolutionary algorithms are that kind of applications. In general,
real world problems and some combinatorial optimization problems such
as the MasterMind puzzle \cite{verbatim} do require an amount of
computational resources unavailable to the average researcher. On the
other hand, one clear advantage of volunteer computing frameworks is
its availability: in principle, any one can request volunteers. How
many she is able to gather, on the other hand, is an issue that we
will try to approach in this paper.

Besides, since Amazon started selling EC2 several years ago, reliable
and scalable computing resources are also available for a low price
and on demand. Recently, Google has also refurbished its offering
lowering their prices. This means that the conjunction of free or
low-cost cloud computing engines, volunteer computing systems and
untapped capability of other people's desktop systems can be used for creating
massive, or at least potentially massive, distributed computing
experiments. These experiments can be easily set up using open-source
repository sites like GitHub\footnote{\url{http://www.github.com}} and deployed automatically to Platform as
a Service (PaaS) sites such as
Heroku\footnote{\url{https://www.heroku.com/}} or
OpenShift\footnote{\url{https://www.openshift.com/}} or even
Infrastructure as a Service (IaaS) sites such as Microsoft's Azure or
Amazon's EC2.

@@

Using the web as a resource for distributed, or plainly user based,
evolutionary computation has a long history since JavaScript was
introduced as a browser-based language in 1997 and even before, when
other procedures, including Flash animations, VBScript, ActiveX or Java applets were
used. Java was pointed out in \cite{soares1998get} as a ``language for
internet'', providing some advantages such as multi-architecture compatibility or 
security mechanisms.
%\begin{quote}
%Java can bring some important advantages... solves architectural
%heterogeneity [...] easy to install and [...] built-in security
%mechanisms
%\end{quote} FERGU: quito esta cita y lo pongo sin el quote
% OK, pero ¿por? - JJ
In that work, Soares et al. describe JET, a system that supports
the execution of parallel applications over the Internet. This system has 
a comprehensive statistics support, allowing its use for science,
since it allows the collection of algorithm analytics by the
experiment designer, unlike the other systems compared in that paper.

@@

The server program includes also additional logic for logging to
different files or services, routes to obtain information from the
server, such as the anonymized IPs that have participated in the
experiment, the ordinal number of the experiment that is being performed or the content of the chromosome pool, and code to serve the static pages
and the JavaScript/CSS
code also.

The system described in this paper is fundamentally different from the
one published in 
\cite{arxiv:modeling}. This changes include, besides the inclusion of logging for a
better experiment analysis, the function that checks for the algorithm
termination condition and restarts the chromosome pool when
it does; it includes a configurable termination
condition that checks for the solution and empties the chromosome pool
and logs the end of the experiment. This allows us to run continuously
experiments without restarting the server by hand as we did in the
previous version. The evolutionary algorithm itself has not changed,
but the facilities for modelling user behavior have improved, as shown
later on in the paper.

In this experiment we have used OpenShift (\url{http://openshift.com})
and Heroku (\url{http://heroku.com}) which are Platform as a Service products with a
free tier. It could be also be deployed, with small
or no modification, to other similar PaaS such as BlueMix by IBM,
maybe with a bit of adaptation. Any of them are
free up to a certain level of use and thus fulfill the requirement of
using free resources as a sustainable way of performing distributed
evolutionary computation experiments \cite{Merelo2014,jj:idc:lowcost}.
%
\begin{figure*}[h!tb]
        \centering
        \includegraphics[width=0.9\linewidth]{img/loggly.png}
        \caption{Loggly dashboard showing the evolution of fitness for
        the last 10 minutes of an experiment.}\label{fig:loggly}
\end{figure*}

Programming the server so that it can be deployed, without
modification, to any PaaS, does not add a lot of overhead to the
programming of a framework, but you have to consider
the different capabilities of each one, namely:\begin{itemize}
\item Heroku does not have a permanent filesystem, so if you want to
  create any directory it has to be done when the server is
  started, and it is only guaranteed to last while the application is
  running; this also implies the need for an external logging system
  to keep the results.
\item In order to scale properly, all requests in Openshift are routed
  through a proxy. This means that all requests have, as remote
  address, the proxy's IP, so other ways of obtaining the requests must be found
  out. On the other hand, it does have a permanent directory for
  data. 
\end{itemize}

Every PaaS has its own way of defining environment variables, which is
the mechanism 
used for passing private configuration information such as database
passwords or, in our case, logging system keys, to the
server. This can be done usually through a command-line client ({\tt
  heroku} or {\tt rhc}), which is also used to check the logs or
restart the client. Eventually, we decided to leave the two servers
running.

@@

Some interesting conclusions can be drawn from this data. The first is
not new: we need frameworks able to work with asynchronous algorithms
to take advantage of all the processing power that is available in an
ad-hoc, ephemeral, setup such as the one available in evolutionary
algorithms. The second is that there is going to be CPU available from
volunteers even when no active announcement is made, just from people
that come from a link in a social media page or the Twitter
timeline. However, an announcement brings immediate attention to the
page which will, obviously, depend on the popularity of the social
media and the person announcing it. How this dependency behaves is
still an open issue. Finally it is also an interesting conclusion
that, given enough time and free computing power, we can solve a
difficult optimization problem. Eventually. But for free.

\section{Conclusions}
\label{sec:conc}

@@

It also remains to be done a comparison of single-desktop computer
experiment compared with this parallel setup. It is quite clear that,
due to overhead, several volunteers will be needed to match the speed
of a single multicore desktop computer. This will be explored in
future work.
